---
title: "Lab 10: Distances and the Border Zone"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Melissa May
    email: "melmay@colostate.edu"
format: html
execute:
  echo: true
---

# Lab 10: Distances and the Border Zone

### Loading libraries:

```{r}
# Spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

## Question 1

### 1.1: Define a Projection

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

### 1.2: Get USA State Boundaries

```{r}
usa_states <- aoi_get(state = "conus")

st_crs(usa_states)

usa_states_eqdc <- st_transform(usa_states, crs = eqdc)

st_crs(usa_states_eqdc)
```

### 1.3: Get Country Boundaries for Mexico, the United States of America, and Canada

```{r}
na_countries <- aoi_get(country = c("MX", "CA", "USA"))

st_crs(na_countries)

na_countries_eqdc <- st_transform(na_countries, crs = eqdc)

st_crs(na_countries_eqdc)
```

### 1.4: Get City Locations from the CSV File

```{r}
cities_df <- read_csv("data/uscities.csv")

glimpse(cities_df)
```

```{r}
cities_sf <- st_as_sf(
  cities_df,
  coords = c("lng", "lat"),
  crs = 4326, # WGS84
  remove = FALSE
)

cities_eqdc <- st_transform(cities_sf, crs = eqdc)
```

```{r}
cities_eqdc <- cities_eqdc %>%
  filter(!state_id %in% c("AK", "HI", "PR", "VI", "GU", "MP", "AS"))
```

## Question 2

### 2.1: Distance to USA Border

```{r}
usa_boundary <- st_union(usa_states_eqdc)

usa_border <- st_cast(usa_boundary, "MULTILINESTRING")

cities_eqdc <- cities_eqdc %>%
  mutate(
    dist_to_national_border_m = as.numeric(st_distance(geometry, usa_border)),
    dist_to_national_border_km = dist_to_national_border_m / 1000
  )

farthest_from_border <- cities_eqdc %>%
  select(city, state_id, dist_to_national_border_km) %>%
  arrange(desc(dist_to_national_border_km)) %>%
  head(5) %>%
  st_drop_geometry()

library(flextable)
farthest_table <- flextable(farthest_from_border) %>%
  set_header_labels(
    city = "City",
    state_id = "State",
    dist_to_national_border_km = "Distance to Border (km)"
  ) %>%
  colformat_double(j = "dist_to_national_border_km", digits = 2)

farthest_table
```

### 2.2: Distance to States

```{r}
states_combined <- st_combine(usa_states_eqdc)
state_borders <- st_cast(states_combined, "MULTILINESTRING")

cities_eqdc <- cities_eqdc %>%
  mutate(
    dist_to_state_border_m = as.numeric(st_distance(geometry, state_borders)),
    dist_to_state_border_km = dist_to_state_border_m / 1000
  )

farthest_from_state_border <- cities_eqdc %>%
  select(city, state_id, dist_to_state_border_km) %>%
  arrange(desc(dist_to_state_border_km)) %>%
  head(5) %>%
  st_drop_geometry()

library(flextable)
farthest_state_table <- flextable(farthest_from_state_border) %>%
  set_header_labels(
    city = "City",
    state_id = "State",
    dist_to_state_border_km = "Distance to State Border (km)"
  ) %>%
  colformat_double(j = "dist_to_state_border_km", digits = 2)

farthest_state_table
```

### 2.3: Distance to Mexico

```{r}
mexico <- na_countries_eqdc %>%
  filter(iso_a2 == "MX")

mexico_border <- st_cast(st_geometry(mexico), "MULTILINESTRING")

cities_eqdc <- cities_eqdc %>%
  mutate(
    dist_to_mexico_m = as.numeric(st_distance(geometry, mexico_border)),
    dist_to_mexico_km = dist_to_mexico_m / 1000
  )

farthest_from_mexico <- cities_eqdc %>%
  select(city, state_id, dist_to_mexico_km) %>%
  arrange(desc(dist_to_mexico_km)) %>%
  head(5) %>%
  st_drop_geometry()

mexico_table <- flextable(farthest_from_mexico) %>%
  set_header_labels(
    city = "City",
    state_id = "State",
    dist_to_mexico_km = "Distance to Mexican Border (km)"
  ) %>%
  colformat_double(j = "dist_to_mexico_km", digits = 2)

mexico_table
```

### 2.4: Distance to Canada

```{r}
canada <- na_countries_eqdc %>%
  filter(iso_a2 == "CA")

canada_border <- st_cast(st_geometry(canada), "MULTILINESTRING")

cities_eqdc <- cities_eqdc %>%
  mutate(
    dist_to_canada_m = as.numeric(st_distance(geometry, canada_border)),
    dist_to_canada_km = dist_to_canada_m / 1000
  )

farthest_from_canada <- cities_eqdc %>%
  select(city, state_id, dist_to_canada_km) %>%
  arrange(desc(dist_to_canada_km)) %>%
  head(5) %>%
  st_drop_geometry()

canada_table <- flextable(farthest_from_canada) %>%
  set_header_labels(
    city = "City",
    state_id = "State",
    dist_to_canada_km = "Distance to Canadian Border (km)"
  ) %>%
  colformat_double(j = "dist_to_canada_km", digits = 2)

canada_table
```

## Question 3

### 3.1: Data

```{r}
top_10_cities <- cities_eqdc %>%
  arrange(desc(population)) %>%
  head(10)

ggplot() +
  geom_sf(data = na_countries_eqdc, fill = "lightgray", color = "gray40", size = 0.5) +
  geom_sf(data = st_union(usa_states_eqdc), fill = NA, color = "black", size = 1.5, lty = 1) +
  geom_sf(data = usa_states_eqdc, fill = NA, color = "gray60", size = 0.3, lty = 2) +
  geom_sf(data = top_10_cities, color = "red", size = 3) +
  ggrepel::geom_label_repel(
    data = top_10_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0,
    force = 5,
    box.padding = 0.5,
    segment.color = "black",
    max.overlaps = 15
  ) +
  theme_minimal() +
  labs(
    title = "North America with CONUS States and Top 10 US Cities by Population",
    caption = "Data source: AOI package and uscities.csv"
  ) +
  coord_sf()
```

### 3.2: City Distance from the Border

```{r}
farthest_5_cities <- cities_eqdc %>%
  arrange(desc(dist_to_national_border_km)) %>%
  head(5)

ggplot() +
  geom_sf(data = na_countries_eqdc, fill = "lightgray", color = "gray60", size = 0.2) +
  geom_sf(data = usa_states_eqdc, fill = NA, color = "gray40", size = 0.3) +
  geom_sf(
    data = cities_eqdc,
    aes(color = dist_to_national_border_km),
    size = 0.8,
    alpha = 0.7
  ) +
  geom_sf(
    data = farthest_5_cities,
    color = "red",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = farthest_5_cities,
    aes(label = paste0(city, ", ", state_id, "\n", round(dist_to_national_border_km, 1), " km"), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0,
    force = 3,
    box.padding = 0.5,
    segment.color = "red",
    fill = "white",
    color = "black"
  ) +
  scale_color_viridis_c(
    name = "Distance to\nNational Border (km)",
    option = "plasma"
  ) +
  theme_minimal() +
  labs(
    title = "US Cities by Distance from National Border",
    subtitle = "Highlighting the 5 cities farthest from any border",
    caption = "Data source: uscities.csv and AOI package"
  ) +
  coord_sf()
```

### 3.3: City Distance from Nearest State

```{r}
farthest_state_cities <- cities_eqdc %>%
  arrange(desc(dist_to_state_border_km)) %>%
  head(5)

ggplot() +
  geom_sf(data = na_countries_eqdc, fill = "lightgray", color = "gray60", size = 0.2) +
  geom_sf(data = usa_states_eqdc, fill = NA, color = "black", size = 0.5) +
  geom_sf(
    data = cities_eqdc,
    aes(color = dist_to_state_border_km),
    size = 0.8,
    alpha = 0.7
  ) +
  geom_sf(
    data = farthest_state_cities,
    color = "red",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = farthest_state_cities,
    aes(label = paste0(city, ", ", state_id, "\n", round(dist_to_state_border_km, 1), " km"), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0,
    force = 3,
    box.padding = 0.5,
    segment.color = "red",
    fill = "white",
    color = "black"
  ) +
  scale_color_viridis_c(
    name = "Distance to\nNearest State Border (km)",
    option = "viridis"
  ) +
  theme_minimal() +
  labs(
    title = "US Cities by Distance from Nearest State Border",
    subtitle = "Highlighting the 5 cities farthest from any state border",
    caption = "Data source: uscities.csv and AOI package"
  ) +
  coord_sf()
```

### 3.4: Equidistance boundary from Mexico and Canada

```{r}
cities_eqdc <- cities_eqdc %>%
  mutate(
    dist_diff_mx_ca = abs(dist_to_mexico_km - dist_to_canada_km)
  )

equidistant_cities <- cities_eqdc %>%
  filter(dist_diff_mx_ca <= 100)

top_5_equidistant <- equidistant_cities %>%
  arrange(desc(population)) %>%
  head(5)

ggplot() +
  geom_sf(data = na_countries_eqdc, fill = "lightgray", color = "gray60", size = 0.2) +
  geom_sf(data = usa_states_eqdc, fill = NA, color = "gray40", size = 0.3) +
  geom_sf(data = cities_eqdc, color = "gray70", size = 0.8, alpha = 0.5) +
  gghighlight(dist_diff_mx_ca <= 100, 
              unhighlighted_params = list(color = "gray70", size = 0.5, alpha = 0.2)) +
  geom_sf(
    data = top_5_equidistant,
    color = "red",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = top_5_equidistant,
    aes(label = paste0(city, ", ", state_id, 
                      "\nPop: ", format(population, big.mark = ","),
                      "\nMX: ", round(dist_to_mexico_km, 1), " km",
                      "\nCA: ", round(dist_to_canada_km, 1), " km"), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0,
    force = 3,
    box.padding = 0.5,
    segment.color = "red",
    fill = "white",
    color = "black"
  ) +
  theme_minimal() +
  labs(
    title = "US Cities Equidistant from Mexican and Canadian Borders",
    subtitle = "Cities where the difference between distances to Mexico and Canada is ≤ 100 km",
    caption = "Highlighting the 5 most populous cities in this zone"
  ) +
  coord_sf()
```

## Question 4

### 4.1: Quantifying Border Zone

```{r}
border_zone_km <- 100 * 1.60934

border_zone_cities <- cities_eqdc %>%
  filter(dist_to_national_border_km <= border_zone_km)

total_us_pop <- sum(cities_eqdc$population, na.rm = TRUE)
border_zone_pop <- sum(border_zone_cities$population, na.rm = TRUE)
border_zone_pct <- (border_zone_pop / total_us_pop) * 100

border_zone_stats <- data.frame(
  Metric = c(
    "Number of cities within 100 miles of border",
    "Population living within 100 miles of border",
    "Percentage of total US city population in border zone",
    "ACLU estimate (reference)"
  ),
  Value = c(
    nrow(border_zone_cities),
    border_zone_pop,
    border_zone_pct,
    "About 2/3 of the US population"
  )
)

border_zone_table <- flextable(border_zone_stats) %>%
  set_header_labels(
    Metric = "Metric",
    Value = "Value"
  ) %>%
  colformat_double(j = "Value", digits = 2) %>%
  autofit()

border_zone_table
```

### 4.2: Mapping Border Zone

```{r}
top_10_border_cities <- border_zone_cities %>%
  arrange(desc(population)) %>%
  head(10)

ggplot() +
  geom_sf(data = na_countries_eqdc, fill = "lightgray", color = "gray60", size = 0.2) +
  geom_sf(data = usa_states_eqdc, fill = NA, color = "gray40", size = 0.3) +
  geom_sf(data = cities_eqdc, color = "gray70", size = 0.8, alpha = 0.5) +
  gghighlight(dist_to_national_border_km <= border_zone_km, 
              use_direct_label = FALSE,
              unhighlighted_params = list(color = "gray70", size = 0.5, alpha = 0.2)) +
  scale_color_gradient(
    low = "orange", 
    high = "darkred",
    name = "Population",
    guide = "none"
  ) +
  geom_sf(
    data = top_10_border_cities,
    color = "purple",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = top_10_border_cities,
    aes(label = paste0(city, ", ", state_id, 
                      "\nPop: ", format(population, big.mark = ",")), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0,
    force = 3,
    box.padding = 0.5,
    segment.color = "purple",
    fill = "white",
    color = "black"
  ) +
  theme_minimal() +
  labs(
    title = "US Cities Within the 100-Mile Border Zone",
    subtitle = "Zone where certain Fourth Amendment protections are limited",
    caption = "Highlighting the 10 most populous cities in this zone"
  ) +
  coord_sf()
```

### 4.3: Label Most Populous City in Each State Within the Danger Zone

```{r}
most_populous_by_state <- border_zone_cities %>%
  group_by(state_id) %>%
  slice_max(order_by = population, n = 1) %>%
  ungroup()

ggplot() +
  geom_sf(data = na_countries_eqdc, fill = "lightgray", color = "gray60", size = 0.2) +
  geom_sf(data = usa_states_eqdc, fill = NA, color = "gray40", size = 0.3) +
  geom_sf(data = cities_eqdc, color = "gray70", size = 0.8, alpha = 0.5) +
  gghighlight(dist_to_national_border_km <= border_zone_km, 
              use_direct_label = FALSE,
              unhighlighted_params = list(color = "gray70", size = 0.5, alpha = 0.2)) +
  scale_color_gradient(
    low = "orange", 
    high = "darkred",
    name = "Population",
    guide = "none"
  ) +
  geom_sf(
    data = most_populous_by_state,
    color = "blue",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = most_populous_by_state,
    aes(label = paste0(city, ", ", state_id, 
                      "\nPop: ", format(population, big.mark = ",")), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0,
    force = 3,
    box.padding = 0.5,
    segment.color = "blue",
    fill = "white",
    color = "black"
  ) +
  theme_minimal() +
  labs(
    title = "US Cities Within the 100-Mile Border Zone",
    subtitle = "Most populous city per state in the border zone",
    caption = "Areas where certain Fourth Amendment protections are limited"
  ) +
  coord_sf()
```
